$K$-corrections and filter transformations \\ in the ultraviolet, optical, and near infrared

### 2006-08-01 docs/paper/kcorrect.tex
update docs
git-svn-id: svn+ssh://howdy.physics.nyu.edu/usr/local/svn/kcorrect/trunk@558 75dff457-8751-0410-ab1a-cd701654ce6b

This approach is similar to PCA in that it finds the small spectral
subspace in which galaxies exist, and can in some ways be thought of
as ``nonnegative PCA.''  However, our method has several advantages
over the standard PCA approach. First, the results yield, along with a
subspace in the space of all possible spectra, a natural
interpretation, which is the corresponding subspace in the space of
all possible star-formation histories. Second, it naturally handles
data uncertainties and missing data, which allows it to ignore that
variation which is due purely to statistical errors.  Third, it
handles the complications of observing galaxy spectra photometrically
using broad-band filters of galaxies at varying redshifts, whereas
standard PCA requires a constant grid of observations in rest-frame
wavelength space.

@@

We begin with a basis set of 485 spectral templates. Of these, 450 are
a set of instantaneous bursts from \citet{bruzual03a}, using the
\citet{chabrier03a} stellar initial mass function and the Padova1994
isochrones. We select 6 metallicities (0.005, 0.02, 0.2, 0.4, 1, and
2.5 times solar). For each metallicity we select 25 ages (between 1
Myr and 13.75 Gyr, spaced almost logarithmically in age).  For each
age and metallicity we make 3 choices of dust model: (1) no dust
extinction; (2) $\tau_V = 3$ dust with Milky Way type extinction (as
input into the models of
\citealt{witt00a} with a ``homogeneous'' distribution and ``shell''
geometry); (3) $\tau_V = 3$ dust with SMC type extinction (from the
same models). We smooth each template to $300$ km s$^{-1}$ velocity
dispersion (we will smooth the SDSS spectra in the training set to the
same resolution).

The remaining 35 templates are from MAPPINGS-III (\citealt{kewley01a})
models of emission from ionized gas. We choose the predictions for an
8 Myr old continuous star-formation history with 5 possible
metalliciticites (0.5, 0.2, 0.4, 1.0, and 2. times solar) and 7
possible ionization parameters ($q= 5\times 10^6$, $10^7$, $2\times
10^7$, $4\times 10^7$, $8 \times 10^7$, $1.5\times 10^8$, and $3\times
10^8$ cm s$^{-1}$). We take the spectra (given as a set of delta
functions) and smooth them to $300$ km s$^{-1}$ velocity
dispersion. The one alteration we make to all of these templates is to
remove Ly$\alpha$, because it is generally much larger than observed
in real galaxies.

Let us refer to these 485 basis templates as $M_{j}(\lambda)$,
expressed in units of ergs s$^{-1}$ \AA$^{-1}$.

We seek to reduce this full basis space to a subspace in which
galaxies actually exist. In particular, we seek five templates
$F_{i}(\lambda)$ built from nonnegative combinations of the
original basis set of $N$ templates:
\begin{equation}
F_{i}(\lambda) = \sum_j b_{ij} M_{j}(\lambda),
\end{equation}
in units of ergs s$^{-1}$ \AA$^{-1}$. In principle, we could seek any
number of templates; from our experiments, we have found that five
turns out to be a number large enough to explain the data we use
here. For each galaxy $k$ we want our model
${\hat{F}_\lambda}(\lambda)$ for their spectrum to be a nonegative sum
of these five templates:
\begin{equation}
{\hat{F}}_{k}(\lambda) = \sum_i a_{ki} F_{i}(\lambda).
\end{equation}

@@

The SDSS data processing consists of astrometry \citep{pier03a};
source identification, deblending and photometry \citep{lupton01a};
photometricity determination \citep{hogg01a}; calibration
\citep{fukugita96a,smith02a}; spectroscopic target selection
\citep{eisenstein01a,strauss02a,richards02a}; spectroscopic fiber
placement \citep{blanton03a}; and spectroscopic data reduction.  We
recalibrated our photometry using the ``ubercalibration'' procedure
described in \citet{blanton05a}.  Descriptions of these pipelines also
exist in \citet{stoughton02a}.  An automated pipeline called {\tt
idlspec2d} (in this case {\tt v4\_9}) measures the redshifts and
classifies the reduced spectra (Schlegel et al., in preparation). We
use the Petrosian magnitudes for the SDSS data (except where noted
below for the LRG-only sample).

@@

We begin with the simplest case, which is fitting a {\it single}
template to the photometric data of the LRG sample
(\citealt{eisenstein01a}). For many of these galaxies, which extend to
$r<19.5$ and are intrinsically red, the $u$-band flux is extremely
poorly measured, so we ignore the $u$-band for all LRGs.

Figure \ref{spec_lrg} shows the spectrum of the best-fit LRG.  This
template is constrained by data between about 2000 \AA\ and 10000 \AA;
outside that range it is an extrapolation.

@@

We next apply the method to the full data set and fit for five
templates. From experiments, we found that we needed five to
adequately explain the data. More than five templates would allow us
to explain the data better, but only marginally so.  The right-most
column of Figure \ref{sfh_templates} shows the spectrum from the
ultraviolet through the near infrared for the resulting
templates. Note that there is a very old template, a very young
template, and several intermediate templates, including one which is
close to that of an A star.

@@

These templates explain the photometric data rather well. For example,
consider Figure \ref{fullfits}, showing the color residuals of the
observations with respect to the best fits, when fitting to galaxies
with GALEX, SDSS, and 2MASS data. In this context we define the color
residuals as (for example):
\begin{equation}
\Delta [u-g] = [u_{\mathrm{obs}}- g_{\mathrm{obs}}] -
[u_{\mathrm{model}}- g_{\mathrm{model}}] .
\end{equation}
In the code described in Section \ref{summary}, we used the routine {\tt
gst\_kcorrect} to perform these fits.  In Figure \ref{fullfits} the
thick dashed lines show the estimated 1$\sigma$ uncertainties in the
colors from the photometric catalogs. Relative to the uncertainties,
there are no significant biases or redshift trends in these fits. Note
that the galaxies used in this test are distinct from the training set
on which we fit the five templates.

@@

In addition to $K$-corrections, these template fits also provide
physical interpretations of the galaxies, since they correspond to
actual stellar population synthesis models. Indeed, the code outputs
three parameters relevant to the star-formation history: the current
mass of stars in the galaxy; the stellar mass-to-light ratio of the
galaxy; and the fraction of the total amount of star-formation that
has occurred recently.

@@

@@

The authors would like to thank Alison Coil, Michael Cooper, David
W. Hogg, John Moustakas, Leonidas Moustakas, David Schiminovich, Risa
Wechsler, and Andrew West for their detailed comments and
feedback. This work was funded by a GALEX Archival Program (38) and an
HST Archival Research grant (AR-9912).

>>>>

This approach is similar to PCA in that it finds the small spectral
subspace in which galaxies exist, and can in some ways be thought of
as ``nonnegative PCA.''  However, our method has several advantages
over the standard PCA approach. First, the results yield, along with a
subspace in the space of all possible spectra, a natural
interpretation, which is the corresponding subspace in the space of
all possible star-formation histories. Second, it naturally handles
data uncertainties and missing data, which allows it to ignore that
variation which is due purely to statistical errors. PCA, by
constrast, does not use any information about the known uncertainties
and is therefore left free to assign templates to express variations
that are purely due to errors. Third, it handles the complications of
observing galaxy spectra photometrically using broad-band filters of
galaxies at varying redshifts, whereas standard PCA requires a
constant grid of observations in rest-frame wavelength space.

@@

We begin with a basis set of 485 spectral templates. Of these, 450 are
a set of instantaneous bursts from \citet{bruzual03a}, using the
\citet{chabrier03a} stellar initial mass function and the Padova1994
isochrones. We use all 6 metallicities (mass fractions of elements
heavier than He of $Z= 0.0001$, $0.0004$, $0.004$, $0.008$, $0.02$ and
$0.05$).  For each metallicity we select 25 ages (between 1 Myr and
13.75 Gyr, spaced almost logarithmically in age).  We smooth each
template to $300$ km s$^{-1}$ velocity dispersion (we will smooth the
SDSS spectra in the training set to the same resolution).

There are limitations to all stellar population synthesis libraries,
and our results are therefore restricted to be at best as good as our
input models. There are some well-known areas of uncertainty in the
stellar populations models in the UV and IR. In the UV, old
populations sometimes contain extreme horizontal branch stars that
have lost their envelopes and are thus extremely hot, that dominate
the flux at $\lambda \sim 1500$ \AA. There is certainly no physical
model that predicts when this happens, and no consensus about how
these stars should be accounted for in isochrone synthesis (that is,
what fraction of stars undergo this phase as a function of mass and
metallicity). In the near-IR, the spectra of thermally-pulsating
asymptotic giant branch stars may dominate the flux in certain
intermediate age populations (2--3 Gyr). Again, the fraction of stars
in this phase and what their spectra are is under dispute. {\bf what
does BC03 do?}  We do not address these uncertainties here, though we
note that for low redshift galaxies our results from BC03 do explain
most galaxies' colors (though see below for our discussion of the high
redshift GOODS sample). {\bf refs}

For each age and metallicity we make 3 choices of dust model: (1) no
dust extinction; (2) $\tau_V = 3$ dust with Milky Way type extinction
(as input into the models of \citealt{witt00a} with a ``homogeneous''
distribution and ``shell'' geometry); (3) $\tau_V = 3$ dust with SMC
type extinction (from the same models).

The remaining 35 templates are from MAPPINGS-III (\citealt{kewley01a})
models of emission from ionized gas. We choose the predictions for an
8 Myr old continuous star-formation history with 5 possible
metalliciticites ($Z=0.001$, 0.004, 0.008, 0.02, and 0.04) and 7
possible ionization parameters ($q= 5\times 10^6$, $10^7$, $2\times
10^7$, $4\times 10^7$, $8 \times 10^7$, $1.5\times 10^8$, and $3\times
10^8$ cm s$^{-1}$). We take the spectra (given as a set of delta
functions) and smooth them to $300$ km s$^{-1}$ velocity
dispersion. The one alteration we make to all of these templates is to
remove Ly$\alpha$, because it is generally much larger than observed
in real galaxies. We warn readers that galaxies exhibit many
individual variations in equivalent widths of emission lines --- our
final set of five templates will not have any hope of following these
variations in detail.

Let us refer to these 485 basis templates as $M_{j}(\lambda)$,
expressed in units of ergs s$^{-1}$ \AA$^{-1}$.  We seek to reduce
this full basis space to a subspace in which galaxies actually
exist. In particular, we seek five templates $F_{i}(\lambda)$ built
from nonnegative combinations of the original basis set of $N$
templates:
\begin{equation}
F_{i}(\lambda) = \sum_j b_{ij} M_{j}(\lambda),
\end{equation}
in units of ergs s$^{-1}$ \AA$^{-1}$. In principle, we could seek any
number of templates; from our experiments, we have found that five
turns out to be a number large enough to explain the data we use
here. For each galaxy $k$ we want our model
${\hat{F}_\lambda}(\lambda)$ for their spectrum to be a nonegative sum
of these five templates:
\begin{equation}
{\hat{F}}_{k}(\lambda) = \sum_i a_{ki} F_{i}(\lambda).
\end{equation}

@@

The SDSS data processing consists of astrometry \citep{pier03a};
source identification, deblending and photometry \citep{lupton01a};
photometricity determination \citep{hogg01a}; calibration
\citep{fukugita96a,smith02a}; spectroscopic target selection
\citep{eisenstein01a,strauss02a,richards02a}; spectroscopic fiber
placement \citep{blanton03a}; and spectroscopic data reduction.  We
recalibrated our photometry using the ``ubercalibration'' procedure
described in \citet{blanton05a}.  Descriptions of these pipelines also
exist in \citet{stoughton02a}.  An automated pipeline called {\tt
idlspec2d}\footnote{Available on the World Wide Web at {\tt
http://skymaps.info}} (in this case {\tt v4\_9}) measures the
redshifts and classifies the reduced spectra (Schlegel et al., in
preparation). We use the Petrosian magnitudes for the SDSS data
(except where noted below for the LRG-only sample). Using the measured
velocity dispersions of the spectra from {\tt idlspec2d}, we further
smooth all spectra in the training sets such that each spectrum has a
dispersion of 300 km s$^{-1}$, to ensure consistency of resolution
among all the spectra and models.

@@

We begin with the simplest case, which is fitting a {\it single}
template to just the SDSS photometric data of the LRG sample
(\citealt{eisenstein01a}). For many of these galaxies, which extend to
$r<19.5$ and are intrinsically red, the $u$-band flux is extremely
poorly measured, so we ignore the $u$-band for all LRGs and only fit
to $g$, $r$, $i$ and $z$ fluxes using the SDSS ``model'' fluxes as
described in \cite{stoughton01a} and elsewhere.
In this case, we fit for the $a_{i0}$ and $b_{0j}$ coefficients 
in Equation \ref{chi2}: we want just a single template, expressed 
by the $b_{0j}$ coefficients, and a single scaling of its 
flux for each galaxy, expressed by $a_{i0}$.

Figure \ref{spec_lrg} shows the spectrum of the best-fit LRG.  This
template is constrained by data between about 3000 \AA\ and 10000 \AA.
Outside that range it is an extrapolation --- that is, it is simply
the best fit model prediction, but there is no data to verify the
model there.

@@

We next apply the method to the full data set (that is, all five items
listed in Section \ref{data}) and fit for five templates. From
experiments, we found that we needed five to adequately explain the
data. More than five templates would allow us to explain the data
better, but only marginally so.  The right-most column of Figure
\ref{sfh_templates} shows the spectrum from the ultraviolet through
the near infrared for the resulting templates. Note that there is a
very old template, a very young template, and several intermediate
templates, including one which is close to that of an A star.

@@

These templates explain the photometric data rather well. For example,
consider Figure \ref{fullfits}, showing the color residuals of the
observations with respect to the best fits, when fitting to galaxies
with GALEX, SDSS, and 2MASS data (using analogous sets of data to 
those described in Section \ref{data}, but in detail a different
set of galaxies). In this context we define the color
residuals as (for example):
\begin{equation}
\Delta [u-g] = [u_{\mathrm{obs}}- g_{\mathrm{obs}}] -
[u_{\mathrm{model}}- g_{\mathrm{model}}] .
\end{equation}
In the code described in Section \ref{summary}, we used the routine {\tt
gst\_kcorrect} to perform these fits.  In Figure \ref{fullfits} the
thick dashed lines show the estimated 1$\sigma$ uncertainties in the
colors from the photometric catalogs. Relative to the uncertainties,
there are no significant biases or redshift trends in these fits. Note
that the galaxies used in this test are distinct from the training set
on which we fit the five templates.

@@

In addition to $K$-corrections, these template fits also provide
physical interpretations of the galaxies, since they correspond to
actual stellar population synthesis models. Indeed, the code outputs
three parameters relevant to the star-formation history: the current
mass of stars in the galaxy; the stellar mass-to-light ratio of the
galaxy; and the fraction of the total amount of star-formation that
has occurred recently. We do not believe that fitting just to
broad-band colors can ever give extremely accurate values for these
quantities.  However, in this section

@@

%The strength of the technique we have used here is that one can
%quickly and reliably whittle down a large parameter space of models to
%the few templates one needs to explain a particular set of data. We
%should note that when fitting models, the most difficult part is
%actually starting with an accurate set of models that are capable of
%explaining the data at all. Here we have used the \citet{bruzual03a}
%models, that do a reasonable but not perfect job of explaining the
%available data.

@@

The authors would like to thank Alison Coil, Michael Cooper, David
W. Hogg, John Moustakas, Leonidas Moustakas, David Schiminovich, Risa
Wechsler, Andrew West, and the anonymous referee for their detailed comments and
feedback. This work was funded by a GALEX Archival Program (38) and an
HST Archival Research grant (AR-9912).
